---
title: "Chapter 11 - Joint Analysis of Longitudinal and Survival Data"
---

## Slides

Lecture slides [here](chap11.html){target="_blank"}. (To convert html to pdf, press E $\to$ Print $\to$ Destination: Save to pdf)

## Chapter Summary

When longitudinal biomarkers are repeatedly measured alongside a time-to-event outcome, separate analyses can lead to biased or inefficient results, especially in the presence of informative dropout. *Joint models* address this by integrating a mixed-effects model for the biomarker process with a survival model for the event time, providing a coherent framework for estimation, interpretation, and prediction.

### Longitudinal–survival joint models

The data consist of repeated measurements $Y_{ij}$ over time $t_{ij}$ $(j=1,\ldots, n_i)$ and a survival outcome $T_i$ subject to right censoring.

-   **Linear mixed-effects (LME) model**\
    For continuous biomarkers, we assume:
    $$
    Y_{ij} = \gamma_0 + \gamma^\mathrm{T} Z_{ij} + b_i^\mathrm{T} \tilde{Z}_{ij} + \epsilon_{ij},
    $$
    where $b_i \sim \mathcal{N}(0, \Sigma_b)$ are subject-specific random effects and $\epsilon_{ij} \sim \mathcal{N}(0, \sigma^2)$ are measurement errors.

    -   *Interpretation*: $\gamma$ are population-level effects; $b_i$ captures individual-specific deviation.
    -   *Estimation*: Maximum likelihood via EM algorithm with $b_i$ as missing data.

-   **Cox model with latent trajectory**\
    The survival process is linked to the "true" biomarker trajectory $m_i(t)$:
    $$
    \Pr(t \le T_i < t + \mathrm{d}t \mid Z_i^*, \overline{m}_i(t)) 
    = \lambda_0(t) \exp\left(\beta^\mathrm{T} Z_i^* + \nu m_i(t)\right) \mathrm{d}t.
    $$

    -   $\nu$ quantifies the log-hazard ratio per unit increase in the biomarker.
    -   $m_i(t)$ may be extracted from an LME or generalized LME model.

### Extensions

-   **Effect modification**\
    Interaction of $m_i(t)$ with treatment or baseline factors:
    $$
    \exp\left\{\nu m_i(t) + \tilde{\nu}^\mathrm{T} \tilde{Z}_i^* m_i(t)\right\}.
    $$

-   **Lagged or cumulative effects**\
    Replace $m_i(t)$ with lagged or average values, e.g.,
    $$
    \overline{m}_i(t - \tau_0, t) = \frac{1}{\tau_0} \int_{t - \tau_0}^t m_i(u) \, \mathrm{d}u.
    $$

-   **Rate of change as predictor**\
    Include derivative $m_i'(t)$ in the hazard model:
    $$
    \exp\left\{\nu_0 m_i(t) + \nu_1 m_i'(t)\right\}.
    $$

-   **Non-continuous outcomes**\
    Use generalized linear mixed models (GLMMs):
    $$
    g\left[E\{Y_i(t) \mid Z_i(t), b_i\}\right] = \gamma_0 + \gamma^\mathrm{T} Z_i(t) + b_i^\mathrm{T} \tilde{Z}_i(t).
    $$

-   **Multivariate extensions**\
    For multiple biomarkers and/or multiple events:
    $$
    \Pr(t \le T_{ik} < t + \mathrm{d}t \mid \xi_i, \overline{m}_i(t)) 
    = \xi_i \lambda_{0k}(t) \exp\left(\beta_k^\mathrm{T} Z_i^* + \nu_k^\mathrm{T} m_i(t)\right) \mathrm{d}t.
    $$

### Dynamic prediction



Joint models enable real-time prediction of survival probability or biomarker levels:

-   **Survival probability**:
    $$
    \mathcal{S}_i(t \mid u) = \Pr(T_i > t \mid T_i > u, \overline{Y}_i(u)).
    $$
-   **Future biomarker level**:
    $$
    \mathcal{M}_i(t \mid u) = E[Y_i(t) \mid T_i > u, \overline{Y}_i(u)].
    $$

These are computed via Monte Carlo integration over $p(b_i \mid \overline{Y}_i(u), T_i > u)$.

### Example R code

#### Univariate Gaussian response: `JM` package

Below is a code snippet demonstrating two-stage joint modeling using the `JM` package. Assume a dataset `df` with:

-   `id`: subject ID, `y`: longitudinal response, `obstime`: time of measurement,\
-   `time`: event or censoring time, `status`: event indicator,\
-   covariates: `covariates`.

```{r}
#| eval: false
########################################
# 1. Linear mixed-effects for biomarker
########################################
library(nlme)
longit_sub <- lme(y ~ covariates + obstime,
                  random = ~ obstime | id,
                  data = df)

########################################
# 2. Cox model for survival outcome
########################################
library(survival)
df_surv <- df[!duplicated(df$id), ]
surv_sub <- coxph(Surv(time, status) ~ covariates, 
                  data = df_surv, x = TRUE)

########################################
# 3. Fit joint model
########################################
library(JM)
obj <- jointModel(longit_sub, surv_sub,
                  timeVar = "obstime",
                  method = "piecewise-PH-aGH")
summary(obj)
```

#### More complex outcomes: `JMbayes2` package

Assume longitudinal outcomes `y1` (continuous) and `y2` (binary) with covariates `obstime`, `age`, and `sex`.

```{r}
#| eval: false
########################################
# 1. Fit two longitudinal models
########################################
# y1: continuous outcome (e.g., biomarker)
longit_sub1 <- lme(y1 ~ obstime + age + sex,
                   random = ~ obstime | id, data = df)

# y2: binary outcome (e.g., clinical status)
library(JMbayes2)
longit_sub2 <- mixed_model(y2 ~ obstime + age + sex,
                           random = ~ obstime | id, data = df,
                           family = binomial())

########################################
# 2. Fit survival model
########################################
df_surv <- df[!duplicated(df$id), ]
surv_sub <- coxph(Surv(time, status) ~ age + sex, data = df_surv)

########################################
# 3. Fit joint model with multiple outcomes
########################################
# Define biomarker contributions to hazard
fForms <- list(
  "y1" = ~ value(y1) + slope(y1),
  "y2" = ~ value(y2)
)

obj <- jm(surv_sub, list(longit_sub1, longit_sub2),
          time_var = "obstime",
          functional_forms = fForms)

summary(obj)
```

-   `value(y)`: uses the fitted mean or link-transformed mean (GLMM) as a predictor.
-   `slope(y)`: includes rate of change (i.e., $\mathrm{d} m_i(t)/\mathrm{d} t$).
-   `area(y)`: uses cumulative biomarker value up to time $t$.

`JMbayes2` enables subject-specific predictions of biomarker evolution and survival probabilities using `predict()`:

```{r}
#| eval: false
# Predict for subject with ID = 3, up to time u = 3
newdf <- df[df$id == 3 & df$obstime < 3, ]
newdf$surv_time <- 3
newdf$status <- 0

# Longitudinal prediction
pred_longit <- predict(obj, newdata = newdf, return_newdata = TRUE)

# Survival prediction
pred_surv <- predict(obj, newdata = newdf, process = "event", return_newdata = TRUE)

```

Use `plot()` to visualize predictions across biomarkers and time horizons, including posterior uncertainty bands.

### Conclusion

Joint modeling of longitudinal and survival data integrates biomarker evolution and event risk to improve efficiency, reduce bias from informative dropout, and provide individualized predictions. Extensions accommodate multiple or non-Gaussian biomarkers, flexible risk structures, and varying functional forms. These models are especially valuable in clinical studies where patient monitoring and outcome prediction are central goals.

## R Code

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false

###############################################################################
# Chapter 11 R Code
#
# This script reproduces all major numerical results in Chapter 11, including:
#   1. Joint modeling of HIV/AIDS data (antiretroviral trial) using {JM}
#   2. Multivariate joint modeling for heart valve study using {JMbayes2}
###############################################################################


#==============================================================================
# (A) Analysis of the Antiretroviral Drug Trial
#==============================================================================
library(tidyverse)   # For data manipulation and ggplot
library(survival)    # For survival analysis (e.g., coxph)
library(JM)          # Joint Modeling package

#------------------------------------------------------------------------------
# 1. Read and Prepare Data
#------------------------------------------------------------------------------
df <- read.table("Data//Anti-retroviral Trial//aids.txt")
head(df)

# Create a de-duplicated data set for the survival sub-model
df_surv <- df[!duplicated(df$id), ]

#------------------------------------------------------------------------------
# 2. Longitudinal Sub-Model
#------------------------------------------------------------------------------
# Linear trajectory for sqrt(CD4) over time
longit_sub <- lme(
  CD4 ~ obsmo + obsmo:drug + sex + hist,
  random = ~ obsmo | id,
  data   = df
)

# Alternative with natural splines:
# longit_sub <- lme(
#   CD4 ~ ns(obsmo, 3) + ns(obsmo, 3):drug + sex + hist,
#   random = list(id = pdDiag(form = ~ ns(time, 3))),
#   data   = df
# )

#------------------------------------------------------------------------------
# 3. Survival Sub-Model
#------------------------------------------------------------------------------
surv_sub <- coxph(
  Surv(time, status) ~ drug + sex + hist,
  data = df_surv,
  x    = TRUE
)

#------------------------------------------------------------------------------
# 4. Joint Model (Piecewise-Constant Baseline Hazard)
#------------------------------------------------------------------------------
obj_joint <- jointModel(
  longit_sub,
  surv_sub,
  timeVar = "obsmo",
  method  = "piecewise-PH-aGH"
)

# Alternatively, a 3-month lag:
# obj_join_l3m <- jointModel(
#   longit_sub, surv_sub,
#   timeVar = "obsmo", lag = 3,
#   method  = "piecewise-PH-aGH"
# )

# Print summary
summary(obj_joint)

#------------------------------------------------------------------------------
# 5. Parameter Extraction and Basic Table
#------------------------------------------------------------------------------
coef   <- obj_joint$coefficients   # Coefficients from model
coef$D                             # Covariance matrix of random effects
beta   <- c(coef$gammas, coef$alpha)  # Survival-related parameters
gamma  <- coef$betas                  # Longitudinal parameters

hmat <- obj_joint$Hessian
Var  <- solve(hmat)          # Inverse Hessian => approximate variance matrix
se   <- sqrt(diag(Var))

# Prepare table for survival sub-model
se_beta <- se[str_detect(names(se), "T.")][1:4]  # Extract relevant subset
za      <- qnorm(0.975)

cox_tab <- tibble(
  ` `       = c("ddI v. ddC", "Male v. female",
                "Infection history (No v. yes)",
                "CD4 count (mm^3)"),
  HR       = round(exp(beta), 2),
  CI       = paste0(
    "(",
    round(exp(beta - za * se_beta), 2), ", ",
    round(exp(beta + za * se_beta), 2), ")"
  ),
  `p-value`= scales::pvalue(2 * (1 - pnorm(abs(beta / se_beta))))
)

knitr::kable(cox_tab, align = "c")

#------------------------------------------------------------------------------
# 6. Residual Analysis
#------------------------------------------------------------------------------
# Base R diagnostic plots
par(mfrow = c(2, 2))
plot(obj_joint)  # built-in {JM} diagnostic

# Extract residuals
epsilons <- residuals(obj_joint, process = "Longitudinal", type = "Subject")  # subject-level
Rs       <- residuals(obj_joint, process = "Longitudinal", type = "Marginal") # marginal
mart     <- residuals(obj_joint, process = "Event")                           # event-level

# Fitted m_i(t): subtract subject-specific residual from observed
ms <- df$CD4 - epsilons

# Plot 1: Observed vs. fitted Y
pred_v_obs <- tibble(
  x = ms,
  y = df$CD4
) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(color = "gray20") +
  geom_abline(intercept = 0, slope = 1, linewidth = 0.8, linetype = 2) +
  xlim(c(0, 22)) +
  ylim(c(0, 22)) +
  labs(
    x     = expression("Fitted " ~ hat(m)[i](t[j])),
    y     = expression("Observed " ~ Y[ij]),
    title = "Overall fit of longitudinal model"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12))

pred_v_obs

# Plot 2: QQ plot for measurement error
eps_qq <- tibble(
  epsilons = epsilons
) %>%
  ggplot(aes(sample = epsilons)) +
  stat_qq(color = "gray20") +
  stat_qq_line(linewidth = 0.8, linetype = 2) +
  labs(
    x     = "Normal Quantiles",
    y     = expression("Quantiles of " ~ hat(epsilon)[ij]),
    title = "Normal QQ plot of measurement error"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12))

eps_qq

# Plot 3: Residual vs. observed outcome
eps_v_y <- tibble(
  x = df$CD4,
  y = epsilons
) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(color = "gray20") +
  geom_hline(yintercept = 0, linewidth = 0.8, linetype = 2) +
  labs(
    x     = expression("Observed " ~ Y[ij]),
    y     = expression("Residual " ~ hat(epsilon)[ij]),
    title = "Homoscedasticity of measurement error"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12))

eps_v_y

# Plot 4: Martingale residual vs. fitted biomarker
mart_v_mi <- tibble(
  x = ms,
  y = mart
) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(color = "gray20") +
  geom_smooth(color = "black", se = FALSE) +
  geom_hline(yintercept = 0, linetype = 3) +
  labs(
    x     = expression("Fitted " ~ hat(m)[i](t[j])),
    y     = "Martingale residuals",
    title = "Martingale residual vs. biomarker"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12))

mart_v_mi

# Combine with patchwork
library(patchwork)
(pred_v_obs + eps_qq) / (eps_v_y + mart_v_mi)

# Save figure
# ggsave("images/longit_aids_resids.png", width = 8, height = 8)
# ggsave("images/longit_aids_resids.eps", width = 8, height = 8)

#------------------------------------------------------------------------------
# 7. Model-Based Mean Trajectories (Figure)
#------------------------------------------------------------------------------
# Compute the mean trajectories based on the fitted longitudinal sub-model
t <- (0:180) / 10

g0 <- gamma[1]
g1 <- gamma[2]
g3 <- gamma[4]
g4 <- gamma[5]

# ddC: baseline drug, with/without history
ddC_nonhist <- (g0 + g3 + g1 * t)^2
ddC_hist    <- (g0 + g1 * t)^2

# ddI: alternate drug, with/without history
ddI_nonhist <- (g0 + g3 + (g1 + g4)*t)^2
ddI_hist    <- (g0 + (g1 + g4)*t)^2

# Base R: Two-panel plot
par(mfrow = c(1, 2))

plot(
  t, ddC_nonhist,
  type       = "l",
  frame.plot = FALSE,
  main       = "No previous infection",
  xlim       = c(0, 18),
  ylim       = c(0, 120),
  xlab       = "Time (months)",
  ylab       = "Mean CD4 cell count",
  lwd        = 2
)
lines(t, ddI_nonhist, lty = 3, lwd = 2)

plot(
  t, ddC_hist,
  type       = "l",
  frame.plot = FALSE,
  main       = "Previous infection",
  xlim       = c(0, 18),
  ylim       = c(0, 120),
  xlab       = "Time (months)",
  ylab       = "Mean CD4 cell count",
  lwd        = 2
)
lines(t, ddI_hist, lty = 3, lwd = 2)

# ggplot version
m <- length(t)
tibble(
  t    = rep(t, 4),
  CD4  = c(ddC_nonhist, ddI_nonhist, ddC_hist, ddI_hist),
  trt  = rep(rep(c("ddC", "ddI"), each = m), 2),
  hist = rep(c("No previous infection", "Previous infection"), each = 2*m)
) %>%
  ggplot(aes(x = t, y = CD4, linetype = trt)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ hist) +
  scale_x_continuous(breaks = seq(0, 18, 6)) +
  scale_y_continuous(limits = c(0, 120), breaks = seq(0, 120, 30)) +
  labs(
    x        = "Time (months)",
    y        = expression("CD4 cell count / mm"^3),
    linetype = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    strip.text      = element_text(size = 11),
    legend.text     = element_text(size = 11),
    legend.key.width= unit(1, "cm")
  )

# ggsave("images/longit_mean_cd4.png", width = 8, height = 4.5)
# ggsave("images/longit_mean_cd4.eps", width = 8, height = 4.5)


#==============================================================================
# (B) Multivariate Analysis (JMbayes2) – Heart Valve Study
#==============================================================================
library(JMbayes2)

#------------------------------------------------------------------------------
# 1. Read and Inspect Data
#------------------------------------------------------------------------------
df_valve <- read.table("Data//Heart Valve Implantation Study//heartv.txt", header = TRUE)

# Basic descriptive info
n <- length(unique(df_valve$id))
num_deaths <- df_valve %>%
  group_by(id) %>%
  slice(1) %>%
  filter(status == 1) %>%
  nrow()

#------------------------------------------------------------------------------
# 2. Longitudinal Sub-Models
#------------------------------------------------------------------------------
# (1) Binary outcome: ejection fraction <= 50%
ef_mod <- mixed_model(
  ef50 ~ obsyear + age + sex + vsize + lvef + redo + cabg + acei + hc + emergenc,
  random = ~ obsyear | id,
  data   = df_valve,
  family = binomial()
)
summary(ef_mod)

# (2) Log valve gradient
grad_mod <- lme(
  grad ~ obsyear + age + sex + vsize + lvef + redo + cabg + acei + hc + emergenc,
  random = ~ obsyear | id,
  data   = df_valve
)
summary(grad_mod)

# (3) Log LV mass index
lvmi_mod <- lme(
  lvmi ~ obsyear + age + sex + vsize + lvef + redo + cabg + acei + hc + emergenc,
  random = ~ obsyear | id,
  data   = df_valve
)
summary(lvmi_mod)

#------------------------------------------------------------------------------
# 3. Survival Sub-Model
#------------------------------------------------------------------------------
df_surv_valve <- df_valve[!duplicated(df_valve$id), ]
surv_mod <- coxph(
  Surv(surv_time, status) ~ age + sex,
  data = df_surv_valve
)

#------------------------------------------------------------------------------
# 4. Multivariate Joint Model in JMbayes2
#------------------------------------------------------------------------------
# Define functional forms in the Cox sub-model
fForms <- list(
  "ef50" = ~ value(ef50),
  "grad" = ~ slope(grad) + value(grad),
  "lvmi" = ~ slope(lvmi) + value(lvmi)
)

# Fit the joint model (may require MCMC or default settings)
obj_valve <- jm(
  surv_mod,
  list(ef_mod, grad_mod, lvmi_mod),
  time_var        = "obsyear",
  functional_forms= fForms
)
summary(obj_valve)

#------------------------------------------------------------------------------
# 5. Extract and Tabulate Results
#------------------------------------------------------------------------------
stats <- obj_valve$statistics
stats_mean <- stats$Mean
stats_sd   <- stats$SD

# A) Longitudinal sub-model estimates
gamma_ef50_est <- stats_mean$betas1
gamma_ef50_se  <- stats_sd$betas1

gamma_grad_est <- stats_mean$betas2
gamma_grad_se  <- stats_sd$betas2

gamma_lvmi_est <- stats_mean$betas3
gamma_lvmi_se  <- stats_sd$betas3

za <- qnorm(0.975)

# Helper function to format estimates + CIs
est_ci <- function(est, se, r = 2, exponentiate = TRUE) {
  if (exponentiate) {
    paste0(
      round(exp(est), r), " (",
      round(exp(est - za * se), r), ", ",
      round(exp(est + za * se), r), ")"
    )
  } else {
    paste0(
      round(est, r), " (",
      round(est - za * se, r), ", ",
      round(est + za * se, r), ")"
    )
  }
}

pval_calc <- function(est, se) {
  scales::pvalue(1 - pchisq((est / se)^2, df = 1))
}

# Example table
df_res <- tibble(
  variable = c(
    "(Intercept)", "Years after surgery", "Age (years)", "Male vs. female",
    "Valve size", "LVEF grade (1–3)", "History of surgery", "CABG",
    "ACE inhibitor", "High cholesterol", "Emergency grade (1–3)"
  ),
  EF_OR = est_ci(gamma_ef50_est, gamma_ef50_se, r = 2, exponentiate = TRUE),
  Grad  = est_ci(gamma_grad_est, gamma_grad_se, r = 2, exponentiate = FALSE),
  LVMI  = est_ci(gamma_lvmi_est, gamma_lvmi_se, r = 2, exponentiate = FALSE)
)
df_res

# B) Survival sub-model
nu    <- stats_mean$alphas
nu_se <- stats_sd$alphas

cox_df <- tibble(
  variable = c(
    "EF<= 50% (log-odds)",
    "LV gradient (mmHg)",
    "LV gradient slope (mmHg/yr)",
    "LVMI (g/m^2)",
    "LVMI slope (g/m^2/yr)"
  ),
  HR   = est_ci(nu, nu_se, r = 2, exponentiate = TRUE),
  pval = pval_calc(nu, nu_se)
)
cox_df

#------------------------------------------------------------------------------
# 6. Dynamic Predictions
#------------------------------------------------------------------------------
# Example: time t0=3 years for patient id=3
t0    <- 3
newdf <- df_valve[df_valve$id == 3, ]            # data for this patient
newdf <- newdf[newdf$obsyear < t0, ]             # only times before t0
newdf$status     <- 0                            # assume alive at t0
newdf$surv_time  <- t0

# Predict longitudinal trajectories from t0 to 10
pred_longit <- predict(
  obj_valve,
  newdata     = newdf,
  times       = seq(t0, 10, length.out = 10),
  return_newdata = TRUE
)

# Base R plot
par(mfrow = c(1, 3))
par(mar   = c(5, 5.2, 2, 1))

plot(
  pred_longit,
  outcomes  = 1:3,
  ylab_long = c(
    "Probability of reduced LVEF",
    "Valve gradient (mmHg)",
    expression("LVMI (" ~ g/m^2 ~ ")")
  ),
  xlab        = "Time (years)",
  xlim        = c(0, 10),
  cex_axis    = 1.5,
  cex_xlab    = 1.8,
  cex_ylab_long = 1.8,
  col_points   = "gray20",
  col_line_long= "black",
  fill_CI_long = "gray80"
)

# Predict survival probabilities
pred_surv <- predict(
  obj_valve,
  newdata     = newdf,
  process     = "event",
  times       = seq(t0, 10, length.out = 10),
  return_newdata = TRUE
)

# Combined event + longitudinal
par(mfrow = c(1, 1))
plot(
  pred_longit, pred_surv,
  outcomes   = 1:3,
  fun_event  = function(x) 1 - x,  # survival = 1 - CIF
  ylab_event = "Survival Probabilities",
  ylab_long  = c(
    "Probability of reduced LVEF",
    "LV gradient (mmHg)",
    expression("LVMI (" ~ g/m^2 ~ ")")
  ),
  pos_ylab_long = c(0.5, 40, 300),
  col_points    = "gray20",
  col_line_long = "black",
  fill_CI_long  = "gray80",
  col_line_event= "black",
  fill_CI_event = "gray80",
  xlab   = "Time (years)",
  xlim   = c(0, 10),
  cex_axis= 1.2,
  cex_xlab= 1,
  cex_ylab_long = 1.2,
  cex_ylab_event= 1
)

# Print predicted survival at 10 years
pred_surv %>%
  dplyr::filter(surv_time == 10) %>%
  dplyr::select(id, surv_time, pred_CIF, low_CIF, upp_CIF) %>%
  mutate(
    pred  = round(1 - pred_CIF, 3),
    lower = round(1 - upp_CIF, 3),
    upper = round(1 - low_CIF, 3)
  )

```


