---
title: "Chapter 1 - Introduction"
execute: 
  warning: false
  cache: true
bibliography: references.bib
---

### Slides

Lecture slides [here](chap1.html){target="_blank"}. (To convert html to pdf, press E $\to$ Print $\to$ Destination: Save to pdf)

### Base R Code

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false



###########################################################################
# The following code generates Figure 1.2 and Table 1.11 of chapter 1.   ##
###########################################################################

####################################################
# Figure 1.2                   
# Proper and improper estimates of survival function
####################################################

## read in the GBC mortality data
data <- read.table("Data//German Breast Cancer Study//gbc_mort.txt")

head(data)

# load the survival package needed for Kaplan--Meier curves
library(survival)


#subset data by hormone==1, 2
data1 <- data[data$hormone==1,]
data2 <- data[data$hormone==2,]

# fit group-specific Kaplan--Meier curves
KMfit1 <- survfit(Surv(time,status)~1,data=data1)
KMfit2 <- survfit(Surv(time,status)~1,data=data2)


# Define a function to calculate empirical survival curve
# Argument: x= a vector containing a sample of real numbers
# Output:   t= order unique values of x
#           S =survival probability as a function of t
emp.surv <- function(x){
  n <- length(x)
  t <- sort(unique(x))
  m <- length(t)
  S <- rep(NA,m)
  for (i in 1:m){
    S[i] <- sum(x>t[i])/n
  }
  return(list(t=t,S=S))
}


## event imputation##

obj1.imp <- emp.surv(data1$time)
obj2.imp <- emp.surv(data2$time)

## complete case
obj1.cc <- emp.surv(data1$time[data1$status==1])
obj2.cc <- emp.surv(data2$time[data2$status==1])


## plot KM, event imputed, and complete-case survival curves
par(mfrow=c(1,2))

plot(KMfit1,conf.int=F,cex.axis=1.5,cex.lab=1.5,cex.main=1.5,
     xlab="Time (months)",ylab="Survival rate",
     main="No Hormone",lwd=2)
lines(obj1.imp$t,obj1.imp$S,lty=2,lwd=2)
lines(obj1.cc$t,obj1.cc$S,lty=3,lwd=2)

plot(KMfit2,conf.int=F,cex.axis=1.5,cex.lab=1.5,cex.main=1.5,
     xlab="Time (months)",ylab="Survival rate",
     main="Hormone",lwd=2)
lines(obj2.imp$t,obj2.imp$S,lty=2,lwd=2)
lines(obj2.cc$t,obj2.cc$S,lty=3,lwd=2)

############################################################################



###########################################################
# Table 1.11. Patient characteristics for the GBC study
###########################################################


##A function calculating median (IQR) by 
## binary group
## Input: y=quantitative variable
##        trt=binary group variable
##        decp=number of decimal points
## Output: a row vector containing median (IQR)
##    by the two levels of trt and overall 

Mean.IQR.by.trt <- function(y,trt,decp=1){
  groups <- sort(unique(trt))
  all <- quantile(y)
  g1 <- quantile(y[trt==groups[1]])
  g2 <- quantile(y[trt==groups[2]])
  
  result <- matrix(NA,1,3)
  colnames(result) <- c(groups,"Overall")
  result[1,1] <- paste0(round(g1[3],decp)," (",round(g1[2],decp),", ",round(g1[4],decp),")")
  result[1,2] <- paste0(round(g2[3],decp)," (",round(g2[2],decp),", ",round(g2[4],decp),")")
  result[1,3] <- paste0(round(all[3],decp)," (",round(all[2],decp),", ",round(all[4],decp),")")
  return(result)
}


##A function calculating N (%) by 
## binary group
## Input: x=categorical variable with p levels
##        trt=binary group variable
##        decp=number of decimal points of %
## Output: a p x 3 matrix containing N (%) for each level of x
##    by the two levels of trt and overall 
N.prct.by.trt <- function(x,trt,decp=1){
  groups <- sort(unique(trt))
  x.levels <- sort(unique(x))
  p <- length(x.levels)
  n <- length(x)
  n1 <- length(x[trt==groups[1]])
  n2 <- length(x[trt==groups[2]])
  
  result <- matrix(NA,p,3)
  colnames(result) <- c(groups,"Overall")
  rownames(result) <- x.levels
  
  for (i in 1:p){
    n1i <- sum(x[trt==groups[1]]==x.levels[i])
    n2i <- sum(x[trt==groups[2]]==x.levels[i])
    ni <- sum(x==x.levels[i])
    
    
  result[i,1] <- paste0(n1i," (",round(n1i/n1*100,decp),"%)")
  result[i,2] <- paste0(n2i," (",round(n2i/n2*100,decp),"%)")
  result[i,3] <- paste0(ni," (",round(ni/n*100,decp),"%)")
  }
  
  
  return(result)
}


##Baseline characteristics by hormonal group:
## age (years), menopause, tumor size (mm), tumor grade
## number of nodes, progesterone receptors content (fmol/mg),
## and estrogen receptors content (fmol/mg)
table1 <- rbind(
  Mean.IQR.by.trt(y=data$age,trt=data$hormone),
  N.prct.by.trt(x=data$meno,trt=data$hormone),
  Mean.IQR.by.trt(y=data$size,trt=data$hormone),
  N.prct.by.trt(x=data$grade,trt=data$hormone),
  Mean.IQR.by.trt(y=data$nodes,trt=data$hormone),
  Mean.IQR.by.trt(y=data$prog,trt=data$hormone),
  Mean.IQR.by.trt(y=data$estrg,trt=data$hormone)
  
)


noquote(table1)

#########################################
# Calculate the event rates (per year)  #
#########################################

#############################
#  Death rate               #
#############################


# Numerator: total # of events
# (N of events is sum of status variable)
num.D <- c(sum(data$status[data$hormone==1]), 
        sum(data$status[data$hormone==2]), 
        sum(data$status))

# Denominator: total length of follow-up (year)
denom.D <- c(sum(data$time[data$hormone==1]), 
        sum(data$time[data$hormone==2]), 
        sum(data$time))/12

# Death rate
round(num.D/denom.D,3)

#############################
# Composite event rate      #
#############################

# Read in the complete data
gbc <- read.table("Data//German Breast Cancer Study//gbc.txt")

# subset to first-event data
# Sort the data by time within each id
o <- order(gbc$id,gbc$time)
gbc <- gbc[o,]
#get the first row for each id
data.CE <- gbc[!duplicated(gbc$id),]

# Numerator: total # of events
# an event means (status>0)
num.CE <- c(sum(data.CE$status[data.CE$hormone==1]>0), 
        sum(data.CE$status[data.CE$hormone==2]>0), 
        sum(data.CE$status>0))

# Denominator: total length of follow-up (year)
denom.CE <- c(sum(data.CE$time[data.CE$hormone==1]), 
          sum(data.CE$time[data.CE$hormone==2]), 
          sum(data.CE$time))/12


#CE rate
round(num.CE/denom.CE,3)

```

### Tidyverse Solutions

First load the required packages:

```{r}
library(tidyverse)
library(survival)

```

#### Parsing censored data

Instead of `(time, status)`, sometimes the observed data are stored in a single column with censored observations indicated with a "+", or "\>" sign.

For example, Table 1.1 of @klein2003 lists the times (in months) to relapse of leukemia in the treatment group (6-MP):

```{r}

MP <- c(10, 7, "32+", 23, 22, 6, 16, "34+", "32+", "25+", "11+", "20+", 
        "19+", 6, "17+", "35+", 6, 13, "9+", "6+", "10+")

```

To convert character strings to `(time, status)`, use `parse_number()` to parse out the number and `str_detect()` to detect whether the string contains "+":

```{r}
#| output: false

df <- tibble(
  MP = MP, # for comparison with newly created variables
  time = parse_number(MP), # extract the number
  status = 1 - str_detect(MP, "\\+") # "\\" to escape literal "+"
)

df
# A tibble: 21 × 3
#    MP     time status
#    <chr> <dbl>  <dbl>
#  1 10       10      1
#  2 7         7      1
#  3 32+      32      0
#  4 23       23      1
#  5 22       22      1
#  6 6         6      1
#  7 16       16      1
#  8 34+      34      0
#  9 32+      32      0
# 10 25+      25      0
# ℹ 11 more rows
# ℹ Use `print(n = ...)` to see more rows
```

We can then use this dataset with standard functions for survival analysis,
e.g., plot the Kaplan--Meier curve:

```{r}

km <- survfit(Surv(time, status) ~ 1, df)
plot(km, main = "Relapse of leukemia in 6-MP group", conf.int = FALSE,
     xlab = "Time (months)", ylab = "Relapse-free probabilities", frame = FALSE)

```


#### Facet plotting Fig. 1.2
Fig. 1.2 of the lecture notes overlays different types of survival estimates
by hormonal treatment status in two panels. This can be done by facet plotting
in ggplot2.

First read in the data:
```{r}
#| output: false

## read in the GBC mortality data
data <- read.table("Data//German Breast Cancer Study//gbc_mort.txt")

head(data)
#   id      time status hormone age meno size grade nodes prog estrg
# 1  1 74.819672      0       1  38    1   18     3     5  141   105
# 2  2 65.770492      0       1  52    1   20     1     1   78    14
# 3  3 47.737705      1       1  47    1   30     2     1  422    89
# 4  4  4.852459      0       1  40    1   24     1     3   25    11
# 5  5 61.081967      0       2  64    2   19     2     1   19     9
# 6  6 63.377049      0       2  49    2   56     1     3  356    64

```

The compute the different estimates within each level of `hormone`. Do this 
by using `group_by(hormone)` and performing the needed calculations within
`reframe()`:


```{r}

# Kaplan--Meier estimates
obj <- summary(survfit(Surv(time, status) ~ hormone, data))
## extract the numbers 
km <- tibble(
  t = obj$time,
  surv = obj$surv,
  hormone = parse_number(as.character(obj$strata)), # 1 or 2
  type = "Kaplan-Meier"
) |> 
  add_row( # need to add starting points for each group
    # so the graph starts from (0, 1):
    t = c(0, 0),
    surv = c(1, 1),
    hormone = 1:2,
    type = "Kaplan-Meier"
  )


# empirical survival functions
## event imputation
## empirical survival functions for all times
imp <- data |> 
  group_by(hormone) |> 
  reframe(
    t = time,
    surv = 1- ecdf(time)(time)
  ) |> 
  arrange(t) |> # order by time
  mutate(
    type = "Event-imputation"
  )

## complete case
## empirical survival functions for times with
## status = 1
cc <- data |> 
  filter(status == 1) |> 
  group_by(hormone) |> 
  reframe(
    t = time,
    surv = 1- ecdf(time)(time)
  ) |> 
  arrange(t) |> # order by time
  mutate(
    type = "Complete-case"
  )
```


Now plot the figure:
```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: center

# create a vector to label the panels
hormone_labeller <- c("1" = "No Hormone", "2" = "Hormone")

# creat the plot
km |> 
  add_row(imp) |> 
  add_row(cc) |> # stack the three estimates together
  mutate(type = fct(type)) |> # keep the order of the levels
  ggplot(aes(x = t, y = surv, linetype = type)) +
  geom_step() + # step function
  facet_wrap( ~ hormone, labeller = labeller(hormone = hormone_labeller)) +
                # different panels by hormone status
  theme_bw() +  # black-white theme
  scale_x_continuous(name = "Time (months)", breaks = seq(0, 72, by = 12), 
                     limits = c(0, 72)) +
  scale_y_continuous(name = "Survival rate", limits = c(0, 1)) +
  scale_linetype_manual(name = "Estimator", values = 1:3) + 
                # set the title and values of line types
  theme(
    legend.position = "bottom"
  )
  
```
Does it look the same as Fig. 1.2?

#### Table one


