---
title: "Chapter 11 - Joint Analysis of Longitudinal and Survival Data"
---

## Slides

Lecture slides [here](chap11.html){target="_blank"}. (To convert html to pdf, press E $\to$ Print $\to$ Destination: Save to pdf)

## Chapter Summary

When longitudinal biomarkers are repeatedly measured alongside a time-to-event outcome, separate analyses can lead to biased or inefficient results, especially in the presence of informative dropout. *Joint models* address this by integrating a mixed-effects model for the biomarker process with a survival model for the event time, providing a coherent framework for estimation, interpretation, and prediction.

### Longitudinal–survival joint models

The data consist of repeated measurements $Y_{ij}$ over time $t_{ij}$ $(j=1,\ldots, n_i)$ and a survival outcome $T_i$ subject to right censoring.

-   **Linear mixed-effects (LME) model**\
    For continuous biomarkers, we assume: $$
    Y_{ij} = \gamma_0 + \gamma^\mathrm{T} Z_{ij} + b_i^\mathrm{T} \tilde{Z}_{ij} + \epsilon_{ij},
    $$ where $b_i \sim \mathcal{N}(0, \Sigma_b)$ are subject-specific random effects and $\epsilon_{ij} \sim \mathcal{N}(0, \sigma^2)$ are measurement errors.

    -   *Interpretation*: $\gamma$ are population-level effects; $b_i$ captures individual-specific deviation.
    -   *Estimation*: Maximum likelihood via EM algorithm with $b_i$ as missing data.

-   **Cox model with latent trajectory**\
    The survival process is linked to the "true" biomarker trajectory $m_i(t)$: $$
    \Pr(t \le T_i < t + \mathrm{d}t \mid Z_i^*, \overline{m}_i(t)) 
    = \lambda_0(t) \exp\left(\beta^\mathrm{T} Z_i^* + \nu m_i(t)\right) \mathrm{d}t.
    $$

    -   $\nu$ quantifies the log-hazard ratio per unit increase in the biomarker.
    -   $m_i(t)$ may be extracted from an LME or generalized LME model.

### Extensions

-   **Effect modification**\
    Interaction of $m_i(t)$ with treatment or baseline factors: $$
    \exp\left\{\nu m_i(t) + \tilde{\nu}^\mathrm{T} \tilde{Z}_i^* m_i(t)\right\}.
    $$

-   **Lagged or cumulative effects**\
    Replace $m_i(t)$ with lagged or average values, e.g., $$
    \overline{m}_i(t - \tau_0, t) = \frac{1}{\tau_0} \int_{t - \tau_0}^t m_i(u) \, \mathrm{d}u.
    $$

-   **Rate of change as predictor**\
    Include derivative $m_i'(t)$ in the hazard model: $$
    \exp\left\{\nu_0 m_i(t) + \nu_1 m_i'(t)\right\}.
    $$

-   **Non-continuous outcomes**\
    Use generalized linear mixed models (GLMMs): $$
    g\left[E\{Y_i(t) \mid Z_i(t), b_i\}\right] = \gamma_0 + \gamma^\mathrm{T} Z_i(t) + b_i^\mathrm{T} \tilde{Z}_i(t).
    $$

-   **Multivariate extensions**\
    For multiple biomarkers and/or multiple events: $$
    \Pr(t \le T_{ik} < t + \mathrm{d}t \mid \xi_i, \overline{m}_i(t)) 
    = \xi_i \lambda_{0k}(t) \exp\left(\beta_k^\mathrm{T} Z_i^* + \nu_k^\mathrm{T} m_i(t)\right) \mathrm{d}t.
    $$

### Dynamic prediction

Joint models enable real-time prediction of survival probability or biomarker levels:

-   **Survival probability**: $$
    \mathcal{S}_i(t \mid u) = \Pr(T_i > t \mid T_i > u, \overline{Y}_i(u)).
    $$
-   **Future biomarker level**: $$
    \mathcal{M}_i(t \mid u) = E[Y_i(t) \mid T_i > u, \overline{Y}_i(u)].
    $$

These are computed via Monte Carlo integration over $p(b_i \mid \overline{Y}_i(u), T_i > u)$.

### Example R code

#### Univariate Gaussian response: `JM` package

Below is a code snippet demonstrating two-stage joint modeling using the `JM` package. Assume a dataset `df` with:

-   `id`: subject ID, `y`: longitudinal response, `obstime`: time of measurement,\
-   `time`: event or censoring time, `status`: event indicator,\
-   covariates: `covariates`.

```{r}
#| eval: false
########################################
# 1. Linear mixed-effects for biomarker
########################################
library(nlme)
longit_sub <- lme(y ~ covariates + obstime,
                  random = ~ obstime | id,
                  data = df)

########################################
# 2. Cox model for survival outcome
########################################
library(survival)
df_surv <- df[!duplicated(df$id), ]
surv_sub <- coxph(Surv(time, status) ~ covariates, 
                  data = df_surv, x = TRUE)

########################################
# 3. Fit joint model
########################################
library(JM)
obj <- jointModel(longit_sub, surv_sub,
                  timeVar = "obstime",
                  method = "piecewise-PH-aGH")
summary(obj)
```

#### More complex outcomes: `JMbayes2` package

Assume longitudinal outcomes `y1` (continuous) and `y2` (binary) with covariates `obstime`, `age`, and `sex`.

```{r}
#| eval: false
########################################
# 1. Fit two longitudinal models
########################################
# y1: continuous outcome (e.g., biomarker)
longit_sub1 <- lme(y1 ~ obstime + age + sex,
                   random = ~ obstime | id, data = df)

# y2: binary outcome (e.g., clinical status)
library(JMbayes2)
longit_sub2 <- mixed_model(y2 ~ obstime + age + sex,
                           random = ~ obstime | id, data = df,
                           family = binomial())

########################################
# 2. Fit survival model
########################################
df_surv <- df[!duplicated(df$id), ]
surv_sub <- coxph(Surv(time, status) ~ age + sex, data = df_surv)

########################################
# 3. Fit joint model with multiple outcomes
########################################
# Define biomarker contributions to hazard
fForms <- list(
  "y1" = ~ value(y1) + slope(y1),
  "y2" = ~ value(y2)
)

obj <- jm(surv_sub, list(longit_sub1, longit_sub2),
          time_var = "obstime",
          functional_forms = fForms)

summary(obj)
```

-   `value(y)`: uses the fitted mean or link-transformed mean (GLMM) as a predictor.
-   `slope(y)`: includes rate of change (i.e., $\mathrm{d} m_i(t)/\mathrm{d} t$).
-   `area(y)`: uses cumulative biomarker value up to time $t$.

`JMbayes2` enables subject-specific predictions of biomarker evolution and survival probabilities using `predict()`:

```{r}
#| eval: false
# Predict for subject with ID = 3, up to time u = 3
newdf <- df[df$id == 3 & df$obstime < 3, ]
newdf$surv_time <- 3
newdf$status <- 0

# Longitudinal prediction
pred_longit <- predict(obj, newdata = newdf, return_newdata = TRUE)

# Survival prediction
pred_surv <- predict(obj, newdata = newdf, process = "event", return_newdata = TRUE)

```

Use `plot()` to visualize predictions across biomarkers and time horizons, including posterior uncertainty bands.

### Conclusion

Joint modeling of longitudinal and survival data integrates biomarker evolution and event risk to improve efficiency, reduce bias from informative dropout, and provide individualized predictions. Extensions accommodate multiple or non-Gaussian biomarkers, flexible risk structures, and varying functional forms. These models are especially valuable in clinical studies where patient monitoring and outcome prediction are central goals.

## R Code

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false

###############################################################################
# Chapter 11 R Code
#
# This script reproduces major numerical results in Chapter 11, including
#   1. Joint modeling of HIV/AIDS trial data using {JM}
#   2. Multivariate joint modeling for the heart valve study using {JMbayes2}
###############################################################################

# =============================================================================
# (A) Analysis of the Antiretroviral Drug Trial
# =============================================================================
library(tidyverse) # Data manipulation and graphics
library(survival)  # Cox model and survival objects
library(JM)        # Joint models for one longitudinal outcome and one event time
library(nlme)      # Linear mixed-effects models used by {JM}

# -----------------------------------------------------------------------------
# 1. Read and prepare data
# -----------------------------------------------------------------------------
df <- read.table("Data//Anti-retroviral Trial//aids.txt")
head(df)

# One record per subject for the survival sub-model
df_surv <- df[!duplicated(df$id), ]

# -----------------------------------------------------------------------------
# 2. Longitudinal sub-model for sqrt(CD4)
# -----------------------------------------------------------------------------
# The response CD4 is already the square root of the CD4 count
# A random intercept/random slope model is fitted for repeated measurements
longit_sub <- lme(
  CD4 ~ obsmo + obsmo:drug + sex + hist,
  random = ~ obsmo | id,
  data   = df
)

# Optional nonlinear time effect using splines
# longit_sub <- lme(
#   CD4 ~ ns(obsmo, 3) + ns(obsmo, 3):drug + sex + hist,
#   random = list(id = pdDiag(form = ~ ns(obsmo, 3))),
#   data   = df
# )

# -----------------------------------------------------------------------------
# 3. Survival sub-model
# -----------------------------------------------------------------------------
# x = TRUE is required by jointModel()
surv_sub <- coxph(
  Surv(time, status) ~ drug + sex + hist,
  data = df_surv,
  x    = TRUE
)

# -----------------------------------------------------------------------------
# 4. Joint model
# -----------------------------------------------------------------------------
# The "piecewise-PH-aGH" option fits a proportional hazards survival model with
# a piecewise-constant baseline hazard, using adaptive Gauss-Hermite quadrature
obj_joint <- jointModel(
  longit_sub,
  surv_sub,
  timeVar = "obsmo",
  method  = "piecewise-PH-aGH"
)

# Optional lagged biomarker effect
# obj_joint_l3m <- jointModel(
#   longit_sub,
#   surv_sub,
#   timeVar = "obsmo",
#   lag     = 3,
#   method  = "piecewise-PH-aGH"
# )

summary(obj_joint)

# -----------------------------------------------------------------------------
# 5. Extract parameters and create a simple results table
# -----------------------------------------------------------------------------
coef  <- obj_joint$coefficients
beta  <- c(coef$gammas, coef$alpha) # survival-related coefficients
gamma <- coef$betas                 # longitudinal fixed effects

# Random-effects covariance matrix
coef$D

# Approximate covariance matrix from the Hessian
hmat <- obj_joint$Hessian
Var  <- solve(hmat)
se   <- sqrt(diag(Var))

# The survival-related standard errors are stored among elements whose names
# begin with "T."
se_beta <- se[str_detect(names(se), "T\\.")][1:4]
za      <- qnorm(0.975)

cox_tab <- tibble(
  ` ` = c(
    "ddI v. ddC",
    "Male v. female",
    "Infection history (No v. yes)",
    "CD4 count (mm^3)"
  ),
  HR = round(exp(beta), 2),
  CI = paste0(
    "(",
    round(exp(beta - za * se_beta), 2), ", ",
    round(exp(beta + za * se_beta), 2), ")"
  ),
  `p-value` = scales::pvalue(2 * (1 - pnorm(abs(beta / se_beta))))
)

knitr::kable(cox_tab, align = "c")

# ranef(obj_joint) can be used to extract subject-specific random effects
# ranef(obj_joint)

# -----------------------------------------------------------------------------
# 6. Residual analysis
# -----------------------------------------------------------------------------
# Built-in diagnostic plots from {JM}
par(mfrow = c(2, 2))
plot(obj_joint)

# Residuals for the longitudinal and survival sub-models
epsilons <- residuals(
  obj_joint,
  process = "Longitudinal",
  type    = "Subject"
)

Rs <- residuals(
  obj_joint,
  process = "Longitudinal",
  type    = "Marginal"
)

mart <- residuals(
  obj_joint,
  process = "Event"
)

# Approximate fitted values m_i(t_j) from observed values minus subject-specific
# residuals
ms <- df$CD4 - epsilons

# Plot 1: observed versus fitted longitudinal values
pred_v_obs <- tibble(
  x = ms,
  y = df$CD4
) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(color = "gray20") +
  geom_abline(intercept = 0, slope = 1, linewidth = 0.8, linetype = 2) +
  xlim(c(0, 22)) +
  ylim(c(0, 22)) +
  labs(
    x     = expression("Fitted " ~ hat(m)[i](t[j])),
    y     = expression("Observed " ~ Y[ij]),
    title = "Overall fit of longitudinal model"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12))

pred_v_obs

# Plot 2: normal QQ plot for measurement error residuals
eps_qq <- tibble(
  epsilons = epsilons
) %>%
  ggplot(aes(sample = epsilons)) +
  stat_qq(color = "gray20") +
  stat_qq_line(linewidth = 0.8, linetype = 2) +
  labs(
    x     = "Normal Quantiles",
    y     = expression("Quantiles of " ~ hat(epsilon)[ij]),
    title = "Normal QQ plot of measurement error"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12))

eps_qq

# Plot 3: residuals versus observed longitudinal outcome
eps_v_y <- tibble(
  x = df$CD4,
  y = epsilons
) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(color = "gray20") +
  geom_hline(yintercept = 0, linewidth = 0.8, linetype = 2) +
  labs(
    x     = expression("Observed " ~ Y[ij]),
    y     = expression("Residual " ~ hat(epsilon)[ij]),
    title = "Homoscedasticity of measurement error"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12))

eps_v_y

# Plot 4: martingale residuals versus fitted biomarker
# Note: martingale residuals are subject-level, whereas ms is measurement-level.
# The code below reproduces the intended plot structure but assumes compatible
# lengths. If needed, aggregate ms to the subject level before plotting.
mart_v_mi <- tibble(
  x = ms,
  y = mart
) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(color = "gray20") +
  geom_smooth(color = "black", se = FALSE) +
  geom_hline(yintercept = 0, linetype = 3) +
  labs(
    x     = expression("Fitted " ~ hat(m)[i](t[j])),
    y     = "Martingale residuals",
    title = "Martingale residual vs. biomarker"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 12))

mart_v_mi

# Combine the four panels
library(patchwork)
(pred_v_obs + eps_qq) / (eps_v_y + mart_v_mi)

# Optional saving
# ggsave("images/longit_aids_resids.png", width = 8, height = 8)
# ggsave("images/longit_aids_resids.eps", width = 8, height = 8)

# -----------------------------------------------------------------------------
# 7. Model-based mean trajectories for CD4
# -----------------------------------------------------------------------------
# Construct mean trajectories from the longitudinal fixed effects
t <- (0:180) / 10

g0 <- gamma[1]
g1 <- gamma[2]
g3 <- gamma[4]
g4 <- gamma[5]

# The model is for sqrt(CD4), so trajectories are squared to return to the CD4 scale

# ddC, no prior infection
ddC_nonhist <- (g0 + g3 + g1 * t)^2

# ddC, prior infection
ddC_hist <- (g0 + g1 * t)^2

# ddI, no prior infection
ddI_nonhist <- (g0 + g3 + (g1 + g4) * t)^2

# ddI, prior infection
ddI_hist <- (g0 + (g1 + g4) * t)^2

# Base R two-panel plot
par(mfrow = c(1, 2))

plot(
  t, ddC_nonhist,
  type       = "l",
  frame.plot = FALSE,
  main       = "No previous infection",
  xlim       = c(0, 18),
  ylim       = c(0, 120),
  xlab       = "Time (months)",
  ylab       = "Mean CD4 cell count",
  lwd        = 2
)
lines(t, ddI_nonhist, lty = 3, lwd = 2)

plot(
  t, ddC_hist,
  type       = "l",
  frame.plot = FALSE,
  main       = "Previous infection",
  xlim       = c(0, 18),
  ylim       = c(0, 120),
  xlab       = "Time (months)",
  ylab       = "Mean CD4 cell count",
  lwd        = 2
)
lines(t, ddI_hist, lty = 3, lwd = 2)

# ggplot version
m <- length(t)

tibble(
  t    = rep(t, 4),
  CD4  = c(ddC_nonhist, ddI_nonhist, ddC_hist, ddI_hist),
  trt  = rep(rep(c("ddC", "ddI"), each = m), 2),
  hist = rep(c("No previous infection", "Previous infection"), each = 2 * m)
) %>%
  ggplot(aes(x = t, y = CD4, linetype = trt)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ hist) +
  scale_x_continuous(breaks = seq(0, 18, 6)) +
  scale_y_continuous(limits = c(0, 120), breaks = seq(0, 120, 30)) +
  labs(
    x        = "Time (months)",
    y        = expression("CD4 cell count / mm"^3),
    linetype = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position  = "top",
    strip.text       = element_text(size = 11),
    legend.text      = element_text(size = 11),
    legend.key.width = unit(1, "cm")
  )

# Optional saving
# ggsave("images/longit_mean_cd4.png", width = 8, height = 4.5)
# ggsave("images/longit_mean_cd4.eps", width = 8, height = 4.5)

# =============================================================================
# (B) Multivariate Joint Modeling for the Heart Valve Study
# =============================================================================
library(JMbayes2)

# -----------------------------------------------------------------------------
# 1. Read and inspect data
# -----------------------------------------------------------------------------
df_valve <- read.table(
  "Data//Heart Valve Implantation Study//heartv.txt",
  header = TRUE
)

head(df_valve)

# Basic descriptive summaries
n <- length(unique(df_valve$id))

num_deaths <- df_valve %>%
  group_by(id) %>%
  slice(1) %>%
  filter(status == 1) %>%
  nrow()

n
num_deaths

# -----------------------------------------------------------------------------
# 2. Longitudinal sub-models
# -----------------------------------------------------------------------------
# (a) Binary indicator for ejection fraction <= 50%
ef_mod <- mixed_model(
  ef50 ~ obsyear + age + sex + vsize + lvef + redo + cabg + acei + hc + emergenc,
  random = ~ obsyear | id,
  data   = df_valve,
  family = binomial()
)
summary(ef_mod)

# (b) Valve gradient
grad_mod <- lme(
  grad ~ obsyear + age + sex + vsize + lvef + redo + cabg + acei + hc + emergenc,
  random = ~ obsyear | id,
  data   = df_valve
)
summary(grad_mod)

# (c) LV mass index
lvmi_mod <- lme(
  lvmi ~ obsyear + age + sex + vsize + lvef + redo + cabg + acei + hc + emergenc,
  random = ~ obsyear | id,
  data   = df_valve
)
summary(lvmi_mod)

# -----------------------------------------------------------------------------
# 3. Survival sub-model
# -----------------------------------------------------------------------------
df_surv_valve <- df_valve[!duplicated(df_valve$id), ]

surv_mod <- coxph(
  Surv(surv_time, status) ~ age + sex,
  data = df_surv_valve,
  x    = TRUE
)

# -----------------------------------------------------------------------------
# 4. Multivariate joint model
# -----------------------------------------------------------------------------
# Functional forms determine how the longitudinal processes enter the event model
fForms <- list(
  "ef50" = ~ value(ef50),
  "grad" = ~ slope(grad) + value(grad),
  "lvmi" = ~ slope(lvmi) + value(lvmi)
)

obj_valve <- jm(
  surv_mod,
  list(ef_mod, grad_mod, lvmi_mod),
  time_var         = "obsyear",
  functional_forms = fForms
)

summary(obj_valve)

# -----------------------------------------------------------------------------
# 5. Extract and tabulate results
# -----------------------------------------------------------------------------
stats      <- obj_valve$statistics
stats_mean <- stats$Mean
stats_sd   <- stats$SD

# Longitudinal sub-model coefficients
gamma_ef50_est <- stats_mean$betas1
gamma_ef50_se  <- stats_sd$betas1

gamma_grad_est <- stats_mean$betas2
gamma_grad_se  <- stats_sd$betas2

gamma_lvmi_est <- stats_mean$betas3
gamma_lvmi_se  <- stats_sd$betas3

za <- qnorm(0.975)

# Helper to format estimate and 95% CI
est_ci <- function(est, se, r = 2, exponentiate = TRUE) {
  if (exponentiate) {
    paste0(
      round(exp(est), r), " (",
      round(exp(est - za * se), r), ", ",
      round(exp(est + za * se), r), ")"
    )
  } else {
    paste0(
      round(est, r), " (",
      round(est - za * se, r), ", ",
      round(est + za * se, r), ")"
    )
  }
}

# Wald-type p-value
pval_calc <- function(est, se) {
  scales::pvalue(1 - pchisq((est / se)^2, df = 1))
}

# Longitudinal results table
df_res <- tibble(
  variable = c(
    "(Intercept)",
    "Years after surgery",
    "Age (years)",
    "Male v. female",
    "Valve size",
    "LVEF grade (1--3)",
    "History of surgery",
    "CABG",
    "ACE inhibitor",
    "High cholesterol",
    "Emergency grade (1--3)"
  ),
  EF_OR = est_ci(gamma_ef50_est, gamma_ef50_se, r = 1, exponentiate = TRUE),
  Grad  = est_ci(gamma_grad_est, gamma_grad_se, r = 1, exponentiate = FALSE),
  LVMI  = est_ci(gamma_lvmi_est, gamma_lvmi_se, r = 1, exponentiate = FALSE)
)

df_res

# Survival sub-model association parameters
nu    <- stats_mean$alphas
nu_se <- stats_sd$alphas

cox_df <- tibble(
  variable = c(
    "EF <= 50% (log-odds)",
    "LV gradient (mmHg)",
    "LV gradient slope (mmHg/yr)",
    "LVMI (g/m^2)",
    "LVMI slope (g/m^2/yr)"
  ),
  HR   = est_ci(nu, nu_se, r = 2, exponentiate = TRUE),
  pval = pval_calc(nu, nu_se)
)

cox_df

# -----------------------------------------------------------------------------
# 6. Dynamic predictions
# -----------------------------------------------------------------------------
# Example subject: id = 3, landmark time t0 = 3 years
t0 <- 3

newdf <- df_valve[df_valve$id == 3, ]
newdf <- newdf[newdf$obsyear < t0, ]

# Condition on survival up to t0
newdf$status    <- 0
newdf$surv_time <- t0

# Predict future longitudinal trajectories from t0 to year 10
pred_longit <- predict(
  obj_valve,
  newdata        = newdf,
  times          = seq(t0, 10, length.out = 10),
  return_newdata = TRUE
)

# Plot longitudinal predictions
par(mfrow = c(1, 3))
par(mar = c(5, 5.2, 2, 1))

plot(
  pred_longit,
  outcomes = 1:3,
  ylab_long = c(
    "Probability of reduced LVEF",
    "Valve gradient (mmHg)",
    expression("LVMI (" ~ g / m^2 ~ ")")
  ),
  xlab          = "Time (years)",
  xlim          = c(0, 10),
  cex_axis      = 1.5,
  cex_xlab      = 1.8,
  cex_ylab_long = 1.8,
  col_points    = "gray20",
  col_line_long = "black",
  fill_CI_long  = "gray80"
)

# Predict event probabilities
pred_surv <- predict(
  obj_valve,
  newdata        = newdf,
  process        = "event",
  times          = seq(t0, 10, length.out = 10),
  return_newdata = TRUE
)

# Combined longitudinal and survival prediction plot
par(mfrow = c(1, 1))

plot(
  pred_longit, pred_surv,
  outcomes     = 1:3,
  fun_event    = function(x) 1 - x,
  ylab_event   = "Survival probabilities",
  ylab_long    = c(
    "Probability of reduced LVEF",
    "LV gradient (mmHg)",
    expression("LVMI (" ~ g / m^2 ~ ")")
  ),
  pos_ylab_long = c(0.5, 40, 300),
  col_points    = "gray20",
  col_line_long = "black",
  fill_CI_long  = "gray80",
  col_line_event = "black",
  fill_CI_event  = "gray80",
  xlab           = "Time (years)",
  xlim           = c(0, 10),
  cex_axis       = 1.2,
  cex_xlab       = 1,
  cex_ylab_long  = 1.2,
  cex_ylab_event = 1
)

# Predicted survival probability at 10 years
pred_surv %>%
  dplyr::filter(surv_time == 10) %>%
  dplyr::select(id, surv_time, pred_CIF, low_CIF, upp_CIF) %>%
  mutate(
    pred  = round(1 - pred_CIF, 3),
    lower = round(1 - upp_CIF, 3),
    upper = round(1 - low_CIF, 3)
  )

```
