---
title: "Chapter 10 - Competing and Semi-Competing Risks"
---

## Slides

Lecture slides [here](chap10.html){target="_blank"}. (To convert html to pdf, press E $\to$ Print $\to$ Destination: Save to pdf)

## Chapter Summary

When a subject can experience at most one event among multiple *competing risks* (e.g., distinct causes of death) or a non-terminal event followed by a terminal event (*semi-competing risks*), standard survival analysis techniques need adjustments to account for the terminality of all or a subset of the events. These adjustments typically involve defining, modeling, and interpreting the *cause-specific hazard*, *cumulative incidence*, and/or *sub-distribution hazards* functions for competing risks, along with adaptations to semi-competing scenarios.

### Competing risks

In competing risks, each subject can fail at most once from a single cause. The outcome data thus consist of the failure time $T$ along with failure cause $\Delta=1,\ldots, K$.

-   **Cause-specific hazard (CSH)**\
    $$
    \mathrm{d}\Lambda_k^{\rm c}(t) 
    \;=\; 
    \Pr\bigl(t \le T < t+\mathrm{d}t,\; \Delta=k \,\mid\, T \ge t\bigr),
    $$ i.e., the *conditional* incidence of cause $k$ among survivors of any cause.

    -   *Estimation*: Treat other causes as censoring and fit a standard Cox model.\
    -   *Interpretation*: Covariate effects reflect changes in risk *conditional on no failure so far*.

-   **Cumulative Incidence Function (CIF)**\
    $$
    F_k(t) 
    \;=\; 
    \Pr(T \le t,\;\Delta = k),
    $$ the *marginal* probability of having cause $k$ by time $t$.

    -   *Nonparametric*: Gray’s estimator integrates the cause-specific hazard with overall survival.\
    -   *Testing*: Gray’s log-rank–type tests compare $F_k(t)$ across groups.\
    -   *Sub-distribution hazard*: A transformation of CIF: $\Lambda_k(t)=-\log\{1 - F_k(t)\}$.

    The *Fine--Gray model* specifies a multiplicative structure on this sub-distribution hazard: $$
    \mathrm{d}\Lambda_k(t\mid Z) \;=\; \exp(\beta_k^\mathrm{T} Z)\,\mathrm{d}\Lambda_{k0}(t),
    $$

    -   $\beta_k$: log-sub-distribution hazard ratios associated with unit increases in the covariates.

-   **Relationship between CSH and CIF** (continuous): \begin{align}
     \mathrm{d}\Lambda_k^{\rm c}(t) 
    &\;=\; \frac{ \mathrm{d} F_k(t)}{1 - \sum_{k'=1}^K F_{k'}(t)};\\
    \mathrm{d} F_k(t)  &\;=\;
    \exp\left\{-\sum_{k'=1}^K \Lambda_{k'}^{\rm c}(t)\right\}\mathrm{d}\Lambda_k^{\rm c}(t).
    \end{align}

### Semi-Competing Risks

Semi-competing risks arise when a non-terminal event (e.g., relapse) can only be observed if it occurs before the terminal event (e.g., death). Once the terminal event happens, the non-terminal event cannot.

-   **Two endpoints** $(T, D)$: $T$ = non-terminal, $D$ = terminal, partially ordered by $T \le D$ if $T$ occurs.\
-   **Recurrent events with death**: $N^*(t)$ = number of non-terminal events by time $t$, with $\mathrm{d} N^*(t)\equiv 0$ for $t> D$ (no event after death).

Analyses can be:

-   *Cause-specific*: treat death as censoring for the non-terminal event.\
-   *Sub-distribution*: treat death as a competing risk for the non-terminal event (acknowledging that there is no event after death).\
-   *Extended frameworks*: (multistate or shared frailty) for more detailed joint modeling.

In particular, the *Ghosh-Lin test* is a two-sample test for comparing the cumulative frequency $E\{N^*(t)\}$, accounting for the terminal event. The *proportional cumulative frequency regression* extends this to covariate-adjusted models: $$
E\{N^*(t)\mid Z\}\;=\;\exp(\beta^\mathrm{T} Z)\mu_0(t).
$$

### Example R code

Below are code snippets demonstrating nonparametric estimation and regression for competing (and semi-competing) risks. Assume a data frame `df` with:

-   `time`: observed time to event or censor,\
-   `status`: integer-coded event indicator (0 = censor, 1, 2 for distinct causes),\
-   covariates: `x1`, `x2`, etc.

```{r}
#| eval: false
########################################
# 1. Proportional cause-specific hazards
########################################
library(survival)
# cause 1 vs. not 1
fit_cs <- coxph(Surv(time, status == 1) ~ x1 + x2, data = df)
summary(fit_cs)

########################################
# 2. Nonparametric CIF (Gray’s method)
########################################
# Using 'cmprsk' to estimate and test cumulative incidence
library(cmprsk)
# Analyses all risks encoded in fstatus
obj_cif <- cuminc(ftime = df$time,
                  fstatus = df$status,
                  group   = df$group   # optional grouping
                  )
obj_cif          # prints CIF estimates & Gray's test p-values
plot(obj_cif)    # basic CIF plot

####################################################
# 3. Fine–Gray proportional sub-distribution hazards
####################################################
library(cmprsk)
# Covariate matrix
cov_mat <- model.matrix(~ x1 + x2, data = df)[, -1]
# cause of interest is 1
fit_fg <- crr(ftime = df$time,
              fstatus = df$status,
              cov1     = cov_mat,
              failcode = 1,
              cencode  = 0)
summary(fit_fg)

####################################################
# 4. Ghosh-Lin methods for cumulative frequency
####################################################
devtools::install_github("lmaowisc/rccf2")
library(rccf2)
# Fit the Ghosh-Lin test for recurrent events
# status = 1: recurrent event; 2: death; 0: censoring
obj_rccf <- rccf(id = df$id, # subject ID
                 time = df$time, # event time
                 status = df$status, # event status
                 trt = df$trt # treatment group
                 )

obj_rccf # print results
```

### Conclusion

When analyzing competing risks, the cause-specific approach (treating other causes as censoring) targets the conditional hazard among survivors, while the cumulative incidence or sub-distribution hazard approach (Fine–Gray) provides a marginal interpretation of failure probability in the presence of competing events. For semi-competing risks---where a non-terminal event cannot follow a terminal event---researchers can treat the terminal event as a competing risk for the non-terminal one or use specialized multistate/frailty methods to capture joint behavior. Each method yields valid but different insights: cause-specific hazards focus on event rates among those alive, while cumulative incidence or sub-distribution models offer population-level probabilities.

## R Code

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false

###############################################################################
# Chapter 10 R Code
#
# This script reproduces major numerical results in Chapter 10, including
#   1. Bone marrow transplantation study (competing risks)
#   2. Semi-competing risks in the German Breast Cancer data
#   3. Bladder tumor study (recurrent events with death as a terminal event)
###############################################################################

# =============================================================================
# (A) Bone Marrow Transplantation Study
# =============================================================================
library(survival)  # Cause-specific hazard modeling and Kaplan--Meier/Cox tools
library(cmprsk)    # Fine--Gray regression and Gray's test
library(tidyverse) # Data manipulation and ggplot2-based graphics

# -----------------------------------------------------------------------------
# 1. Read data and perform Gray's test
# -----------------------------------------------------------------------------
cibmtr <- read.table("Data//Bone Marrow Transplantation Study//cibmtr.txt")
head(cibmtr)

# Gray's test compares cumulative incidence functions across donor groups
# rho = 0 gives the standard unweighted version
obj <- cmprsk::cuminc(cibmtr$time, cibmtr$status, cibmtr$donor, rho = 0)

# Test statistics and p-values for each cause
obj$Tests

# -----------------------------------------------------------------------------
# 2. Plot cumulative incidence functions for relapse and TRM
# -----------------------------------------------------------------------------
# donor = 0: sibling donor
# donor = 1: unrelated donor
# cause 1: relapse
# cause 2: treatment-related mortality (TRM)
obj.rlp.sib    <- obj$`0 1` # cause 1 (relapse) for donor 0 (sibling)
obj.rlp.nonsib <- obj$`1 1` # cause 1 (relapse) for donor 1 (unrelated)
obj.trm.sib    <- obj$`0 2` # cause 2 (TRM) for donor 0 (sibling)
obj.trm.nonsib <- obj$`1 2` # cause 2 (TRM) for donor 1 (unrelated)

par(mfrow = c(1, 2))

# Relapse
plot(
  obj.rlp.sib$time, obj.rlp.sib$est,
  type       = "s",
  frame.plot = FALSE,
  main       = "Relapse",
  xlim       = c(0, 120),
  ylim       = c(0, 0.6),
  xlab       = "Time (months)",
  ylab       = "Cumulative incidence",
  lwd        = 2
)
lines(obj.rlp.nonsib$time, obj.rlp.nonsib$est, lty = 3, lwd = 2)

# Treatment-related mortality
plot(
  obj.trm.sib$time, obj.trm.sib$est,
  type       = "s",
  frame.plot = FALSE,
  main       = "Treatment-related mortality",
  xlim       = c(0, 120),
  ylim       = c(0, 0.6),
  xlab       = "Time (months)",
  ylab       = "Cumulative incidence",
  lwd        = 2
)
lines(obj.trm.nonsib$time, obj.trm.nonsib$est, lty = 3, lwd = 2)

# -----------------------------------------------------------------------------
# 3. Fine--Gray versus cause-specific hazard models for relapse
# -----------------------------------------------------------------------------
# Fine--Gray model targets the subdistribution hazard for cause 1
obj.fg <- cmprsk::crr(cibmtr$time, cibmtr$status, cibmtr[, 3:6], failcode = 1)

# Extract coefficients and standard errors
beta.fg <- obj.fg$coef
se.fg   <- sqrt(diag(obj.fg$var))

# Hazard ratio, 95% CI, and Wald p-value
c1 <- round(exp(beta.fg), 2)
c2 <- paste0(
  "(",
  round(exp(beta.fg - 1.96 * se.fg), 2), "-",
  round(exp(beta.fg + 1.96 * se.fg), 2), ")"
)
c3 <- round(1 - pchisq((beta.fg / se.fg)^2, 1), 3)

# Cause-specific hazard model for relapse
obj.csh <- coxph(Surv(time, status == 1) ~ cohort + donor + hist + wait, data = cibmtr)

beta.csh <- obj.csh$coef
se.csh   <- sqrt(diag(obj.csh$var))

c4 <- round(exp(beta.csh), 2)
c5 <- paste0(
  "(",
  round(exp(beta.csh - 1.96 * se.csh), 2), "-",
  round(exp(beta.csh + 1.96 * se.csh), 2), ")"
)
c6 <- round(1 - pchisq((beta.csh / se.csh)^2, 1), 3)

# Columns 1--3: Fine--Gray
# Columns 4--6: cause-specific hazard model
noquote(cbind(c1, c2, c3, c4, c5, c6))

# -----------------------------------------------------------------------------
# 4. Covariate-specific CIF prediction from the Fine--Gray model
# -----------------------------------------------------------------------------
# Example covariate vector z = (1, 0, 1, 0)
z <- c(1, 0, 1, 0)

# Method 1: built-in prediction
obj_pred1 <- predict(obj.fg, z)

# Method 2: manual construction using the baseline subdistribution hazard
beta   <- obj.fg$coef
Lambda <- cumsum(obj.fg$bfitj)
time   <- obj.fg$uftime
cif    <- 1 - exp(-exp(sum(beta * z)) * Lambda)
obj_pred2 <- cbind(time, cif)

# The two predictions should agree closely
cbind(obj_pred1, obj_pred2)

# =============================================================================
# (B) Tidy Competing Risks Analysis
# =============================================================================
library(tidycmprsk) # Tidy competing-risks workflows and regression
library(ggsurvfit)  # CIF plotting with risk tables

# Recode status and donor for tidycmprsk
df <- cibmtr %>%
  dplyr::mutate(
    status = factor(status, levels = c("0", "1", "2"),
                    labels = c("Censored", "Relapse", "TRM")),
    donor  = factor(donor, levels = c("0", "1"),
                    labels = c("Id. sibling", "Unrelated"))
  )

# -----------------------------------------------------------------------------
# 1. Estimate CIFs by donor group
# -----------------------------------------------------------------------------
obj_cif <- tidycmprsk::cuminc(Surv(time, status) ~ donor, data = df)
obj_cif

# Summaries of CIFs at selected times
tbl_surv <- tbl_cuminc(
  x            = obj_cif,
  times        = seq(12, 120, by = 36),
  outcomes     = c("Relapse", "TRM"),
  label_header = "{time} months",
  label        = "Donor"
)
tbl_surv

# CIF plot for both outcomes together
obj_cif %>%
  ggcuminc(outcome = c("Relapse", "TRM")) +
  scale_x_continuous("Time (months)", limits = c(0, 120), breaks = seq(0, 120, 24)) +
  add_risktable(risktable_stats = "n.risk") +
  theme_minimal()

# -----------------------------------------------------------------------------
# 2. Separate CIF plots by outcome
# -----------------------------------------------------------------------------
library(patchwork) # Combine outcome-specific panels
library(ggsci)     # Journal-style color palettes

plot_bm_cif <- function(obj_cif, outcome) {
  p <- ggcuminc(obj_cif, outcome, linetype_aes = TRUE, linewidth = 0.8) +
    add_risktable(
      risktable_stats = "n.risk",
      theme = list(theme_risktable_default())
    ) +
    ggtitle(outcome) +
    scale_y_continuous(
      "Cumulative incidence",
      limits = c(0, 0.6),
      breaks = seq(0, 1, by = 0.1),
      expand = expansion(c(0, 0.005))
    ) +
    scale_x_continuous("Time (months)", limits = c(0, 120), breaks = seq(0, 120, 24)) +
    scale_color_jama() +
    scale_linetype_manual(values = c(2, 1)) +
    theme_classic() +
    theme(
      plot.margin        = margin(0, 0, 0, 0),
      legend.position    = "top",
      legend.key.width   = unit(1, "cm"),
      panel.grid.major.y = element_line(color = "gray80", size = 0.5),
      legend.text        = element_text(size = 10),
      plot.caption       = element_text(size = 10),
      plot.title         = element_text(size = 12)
    )
  p
}

rel_fig <- plot_bm_cif(obj_cif, "Relapse")
trm_fig <- plot_bm_cif(obj_cif, "TRM")

bm_fig <- wrap_plots(
  ggsurvfit_build(rel_fig),
  ggsurvfit_build(trm_fig),
  ncol = 2
)
bm_fig

# Optional saving
# ggsave("images/cmpr_bm.png", bm_fig, width = 8, height = 4.5)
# ggsave("images/cmpr_bm.eps", bm_fig, width = 8, height = 4.5)

# -----------------------------------------------------------------------------
# 3. Fine--Gray regression for relapse and TRM
# -----------------------------------------------------------------------------
obj_rel <- tidycmprsk::crr(
  Surv(time, status) ~ donor + cohort + hist + wait,
  data = df,
  failcode = "Relapse"
)
obj_rel

obj_trm <- tidycmprsk::crr(
  Surv(time, status) ~ donor + cohort + hist + wait,
  data = df,
  failcode = "TRM"
)
obj_trm

# -----------------------------------------------------------------------------
# 4. Proportional subdistribution odds model
# -----------------------------------------------------------------------------
library(timereg) # Alternative semiparametric competing-risks models

obj_po1 <- prop.odds.subdist(
  Event(time, status) ~ donor + cohort + hist + wait,
  data  = df,
  cause = 1
)

obj_po1$gamma
obj_po1$robvar.gamma

# =============================================================================
# (C) Semi-Competing Risks in the German Breast Cancer Data
# =============================================================================
library(gtsummary) # Regression summary tables
library(labelled)  # Variable labels for printing

# -----------------------------------------------------------------------------
# 1. Read and recode GBC data
# -----------------------------------------------------------------------------
gbc <- read.table("Data//German Breast Cancer Study//gbc.txt") %>%
  mutate(
    age40 = (age > 40) + 0,
    grade = factor(grade),
    prog  = prog / 100,
    estrg = estrg / 100
  )

# Human-readable labels for regression tables
var_label(gbc) <- list(
  hormone = "Hormone",
  meno    = "Menopause",
  age40   = "Age 40+",
  grade   = "Tumor grade",
  size    = "Tumor size (mm)",
  nodes   = "Number of positive lymph nodes",
  prog    = "Progesterone (100 fmol/mg)",
  estrg   = "Estrogen (100 fmol/mg)"
)

# -----------------------------------------------------------------------------
# 2. Cox model for death
# -----------------------------------------------------------------------------
# Remove subjects whose observed first event was relapse only
# Keep death and censoring for overall survival from death
gbc_death <- gbc %>%
  filter(status != 1) %>%
  mutate(status = (status > 0))

obj_death <- coxph(
  Surv(time, status) ~ hormone + meno + age40 + size + grade + nodes + prog + estrg,
  data = gbc_death
)

death_tbl <- tbl_regression(obj_death, exponentiate = TRUE)

# -----------------------------------------------------------------------------
# 3. Cause-specific hazard model for relapse
# -----------------------------------------------------------------------------
# Keep one row per subject corresponding to the earliest time
# Treat death as censoring when modeling relapse as the cause of interest
gbc_relc <- gbc %>%
  group_by(id) %>%
  slice_min(time) %>%
  mutate(status = if_else(status == 2, 0, status)) %>%
  slice_max(status) %>%
  ungroup()

obj_relc <- coxph(
  Surv(time, status) ~ hormone + meno + age40 + size + grade + nodes + prog + estrg,
  data = gbc_relc
)

relc_tbl <- tbl_regression(obj_relc, exponentiate = TRUE)

# -----------------------------------------------------------------------------
# 4. Subdistribution hazard model for relapse
# -----------------------------------------------------------------------------
# Keep the first observed event/censoring record for each subject
gbc_tfe <- gbc %>%
  group_by(id) %>%
  slice_min(time) %>%
  slice_max(status) %>%
  mutate(status = factor(status, levels = c("0", "1", "2"))) %>%
  ungroup()

obj_rel_sub <- tidycmprsk::crr(
  Surv(time, status) ~ hormone + meno + age40 + size + grade + nodes + prog + estrg,
  data     = gbc_tfe,
  failcode = "1"
)

rel_sub_tbl <- tbl_regression(obj_rel_sub, exponentiate = TRUE)

# Merge the three regression tables
tbl <- tbl_merge(
  tbls = list(death_tbl, relc_tbl, rel_sub_tbl),
  tab_spanner = c("**Death**", "**Relapse (cause-specific)**", "**Relapse (subdistribution)**")
)

tbl

# =============================================================================
# (D) Bladder Tumor Study
# =============================================================================
# NOTE: Installing from GitHub creates side effects and may not be needed if the
# package is already installed.
devtools::install_github("lmaowisc/rccf2")
library(rccf2)
data(bladder)

# -----------------------------------------------------------------------------
# 1. Descriptive analysis
# -----------------------------------------------------------------------------
# Frequency of each event type
bladder %>%
  count(status)

# Add subject-level maximum follow-up and recode status/treatment for plotting
df <- bladder %>%
  left_join(
    bladder %>% group_by(id) %>% summarize(max_fu = max(time)),
    by = "id"
  ) %>%
  mutate(
    id     = factor(id),
    status = factor(status, levels = c("0", "1", "2")),
    trt    = if_else(trt == 1, "Thiotepa", "Placebo")
  )

# One row per subject for plotting overall follow-up length
dfmax <- df %>%
  group_by(id, trt) %>%
  summarize(max_fu = max(time), .groups = "drop")

# Follow-up display with recurrence and death markers
df %>%
  ggplot(aes(y = reorder(id, max_fu))) +
  geom_linerange(data = dfmax, aes(xmin = 0, xmax = max_fu)) +
  geom_point(aes(x = time, shape = status), size = 2, fill = "white") +
  geom_vline(xintercept = 0, linewidth = 1) +
  scale_shape_manual(
    values = c(23, 15, 19),
    labels = c("Censoring", "Tumor recurrence", "Death")
  ) +
  scale_x_continuous(
    "Time (months)",
    limits = c(0, 65),
    breaks = seq(0, 60, by = 12),
    expand = c(0, 0.5)
  ) +
  scale_y_discrete("Patients") +
  facet_wrap(~ trt, scales = "free") +
  theme_minimal() +
  theme(
    legend.position    = "top",
    legend.title       = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.grid.major.y = element_blank()
  )

# Optional saving
# ggsave("cmpr_blad_fu.pdf", width = 8, height = 8)
# ggsave("cmpr_blad_fu.eps", width = 8, height = 8)

# -----------------------------------------------------------------------------
# 2. Ghosh--Lin two-sample analysis
# -----------------------------------------------------------------------------
id     <- bladder$id
time   <- bladder$time
status <- bladder$status
trt    <- bladder$trt

# rccf() compares cumulative frequency of recurrent events with death as terminal event
obj_rccf <- rccf(id, time, status, trt)

stat   <- obj_rccf$stat
S      <- obj_rccf$S
u1     <- obj_rccf$u1
u2     <- obj_rccf$u2
t_rccf <- obj_rccf$t

# Partial tests
QLR <- stat[1] / sqrt(S[1, 1])
pLR <- 1 - pchisq(QLR^2, 1)

QD  <- stat[2] / sqrt(S[2, 2])
pD  <- 1 - pchisq(QD^2, 1)

# Joint test for recurrence and death together
Q <- t(stat) %*% solve(S) %*% stat
p <- 1 - pchisq(Q, 2)

# -----------------------------------------------------------------------------
# 3. Plot survival from death and cumulative tumor frequency
# -----------------------------------------------------------------------------
par(mfrow = c(1, 2))

# Survival with death as the event
fatal <- bladder[status != 1, ]
obj_sf <- survfit(Surv(time, status > 0) ~ trt, data = fatal)

plot(
  obj_sf,
  lty      = c(1, 3),
  frame    = FALSE,
  xlim     = c(0, 50),
  lwd      = 2,
  cex.lab  = 1.5,
  cex.axis = 1.5,
  xlab     = "Time (months)",
  ylab     = "Survival probabilities"
)
text(35, 0.14, paste0("Log-rank test: p=", round(pD, 3)), cex = 1.2)

# Cumulative tumor frequency by treatment group
stepmu1 <- stepfun(t_rccf, c(0, u1))
stepmu2 <- stepfun(t_rccf, c(0, u2))

plot(
  stepmu2,
  do.points = FALSE,
  xlab      = "Time (months)",
  ylab      = "Cumulative tumor frequency",
  main      = "",
  xlim      = c(0, 60),
  ylim      = c(0, 3),
  frame     = FALSE,
  lwd       = 2,
  cex.lab   = 1.5,
  cex.axis  = 1.5,
  lty       = 3
)
plot(
  stepmu1,
  add       = TRUE,
  lty       = 1,
  do.points = FALSE,
  lwd       = 2
)
text(40, 0.4, paste0("Ghosh-Lin test: p=", round(pLR, 3)), cex = 1.2)

# -----------------------------------------------------------------------------
# 4. Ghosh--Lin proportional cumulative frequency regression
# -----------------------------------------------------------------------------
library(mets) # Regression for recurrent events with terminal events
data(bladder)

# Create start-stop representation for each subject
df_gl <- bladder %>%
  group_by(id) %>%
  mutate(
    start = lag(time, default = 0),
    stop  = if_else(time == 0, 1e-4, time)
  ) %>%
  select(id, start, stop, status, trt) %>%
  ungroup() %>%
  as.data.frame()

# Fit proportional cumulative frequency regression
obj_reg <- recreg(
  Event(start, stop, status) ~ trt + cluster(id),
  cause      = 1,
  death.code = 2,
  data       = df_gl
)

summary(obj_reg)

```
