---
title: "10-Minute In-Class Quiz 3: Survival Analysis (Chapters 8-9)"
format: pdf
execute:
  echo: false
  warning: false
---

## **Instructions**

-   You have **10 minutes** to complete this quiz.
-   Answer all questions *concisely*.
-   Show all relevant calculations where applicable.

## **Question 1 (5 points) – Recurrent Events**

@fig-followup below shows the follow-up of six patients for repeated infections.

```{r}
#| label: fig-followup
#| fig-cap: Follow-up of six patients for repeated infections.
#| fig-width: 8
#| fig-height: 3

library(tidyverse)

pat1 <- tibble(
 id = 1,
 time = c(0, 1, 3, 4, 5, 6, 8.2),
 status = c(NA, rep(1, 5), 0)
)
pat2 <- tibble(
 id = 2,
 time = c(0, 1, 2, 3.2),
 status = c(NA, rep(1, 2), 0)
)
pat3 <- tibble(
 id = 3,
 time = c(0, 2, 3, 5, 6.2),
 status = c(NA, rep(1, 3), 0)
)
pat4 <- tibble(
 id = 4,
 time = c(0, 1, 2, 3, 4.2),
 status = c(NA, rep(1, 3), 0)
)
pat5 <- tibble(
 id = 5,
 time = c(0, 1.2),
 status = c(NA, 0)
)
pat6 <- tibble(
 id = 6,
 time = c(0, 1, 2, 3, 4, 5, 6, 7.2),
 status = c(NA, rep(1, 6), 0)
)


df <- bind_rows(pat1, pat2, pat3, pat4, pat5, pat6)

df |> 
  ggplot(aes(x = time, y = factor(id))) +
  geom_line(aes(group = id)) +
  geom_point(aes(shape = factor(status)), size = 3, fill = "white") +
  geom_vline(xintercept = 0, linewidth = 1.2) +
  scale_shape_manual(limits = c("0", "1"),  values = c(23, 19), 
                      labels = c("Censoring", "Infection"))  +
  scale_x_continuous(expand = expansion(c(0, 0.02))) +
  labs(
    x = "Time (months)",
    y = "Patient ID",
    shape = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    legend.text = element_text(size = 11)
  )
 
ggsave("images/rec_fu_prob.png", width = 8, height = 3)
ggsave("images/rec_fu_prob.eps", width = 8, height = 3)

```

-   Calculate the (Nelsen--Aalen-type) estimates of the **average number of infections per patient by month 2, 4, and 6**.
    -   `d_j`: observed number of events $d_j$
    -   `n_j`: observed number of patients at risk $n_j$
    -   `cum_avg`: $\sum_{l=1}^j d_{l}/n_l$

```{r}
#| label: tbl-NA-estimates
#| tbl-cap: Nelsen--Aalen-type estimates of the cumulative average number of infections.


library(knitr)
tbl <- tibble(
  month = 1:6,
  d_j = c(4, 4, 4, 2, 3, 2),
  n_j = c(6, 5, 5, 4, 3, 3),
  `d_j/n_j` = round(d_j / n_j, 2),
  cum_avg = round(cumsum(`d_j/n_j`), 2)
)

kable(tbl, align = "c")

```

-   Why is the following arithmetic biased:

$$
\frac{\mbox{Total number of black dots by month }2, 4, 6}{6 \mbox{ subjects}}
$$ 

This is equivalent to setting $n_j=6$ for all months. However, not all six patients are at risk when infections occur. 


## **Question 2 (5 points) – Mean Model**

Let $N^*(t)$ denote the average number of recurrent events by time $t$ (such as estimated in Question 1). Let $Z=1, 0$ denote treatment and control groups, respectively. Under the multiplicative mean model: $$
E\{N^*(t) \mid Z\} = \exp(\beta Z)\mu_0(t),
$$

-   What is the mean function in each group, i.e., $E\{N^*(t) \mid Z = z\}$ for $z= 1, 0$?

    -   $E\{N^*(t) \mid Z = 0\} = \mu_0(t)$;
    -   $E\{N^*(t) \mid Z = 1\} =  \exp(\beta)\mu_0(t)$.


-   Suppose $\hat\beta = \log(0.8)$, interpret the treatment effect.

The treatment group has a **20% reduction** ($1-\exp(\hat\beta)$) in the average number of recurrent events compared to the control group.